---
title: "Car Acceptability Classification"
subtitle: "Exploratory Data Analysis & Machine Learning"
author: "Priya"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	fig.align = "center",
	fig.height = 5,
	fig.width = 7,
	message = FALSE,
	warning = FALSE
)
```

***

# Setup

## Installing & Loading Packages

```{r packages}
packages <- c("nnet", "caret", "class", "gmodels", "C50",
              "klaR", "e1071", "kernlab", "caTools", "ggplot2", "infotheo",
              "dplyr", "reshape2")

new.packages <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(new.packages)) install.packages(new.packages)

invisible(lapply(packages, library, character.only = TRUE))
```

## Importing the Dataset

```{r import-data}
source_file <- "C:/Users/priya/OneDrive - Ulster University/University/CarSelection_Algorithm/car.csv"

df <- read.csv(source_file)
```

***

# Exploratory Data Analysis & Data Preparation

## Initial Inspection

```{r eda-initial}
head(df)

sum(is.na(df))  # Check for missing values

df <- data.frame(lapply(df, factor))  # Convert all columns to factors

summary(df)

str(df)
```

## Counting with `table()`

`table()` lets us cross-tabulate categorical variables and spot patterns quickly.

```{r table-acceptability}
table(df$acceptability)

df$buying <- factor(df$buying, levels = c("vhigh", "high", "med", "low"))

table(df$buying, df$acceptability)
```

> **Insight:** Low-price cars are more commonly rated as acceptable, suggesting price is a key driver of acceptability.

```{r table-safety}
table(df$safety, df$acceptability)
```

> **Insight:** 35 high-safety cars are rated `vgood`, while `low` and `med` safety cars have zero `vgood` ratings — safety is strongly tied to top ratings.

## Visualising Variable Distributions

### Acceptability

```{r plot-acceptability}
df$acceptability <- factor(df$acceptability, levels = c("unacc", "acc", "good", "vgood"))

ggplot(df, aes(x = acceptability)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of Car Acceptability",
       x = "Acceptability", y = "Count") +
  theme_minimal()
```

### Acceptability vs. Buying Price

```{r plot-buying}
ggplot(df, aes(x = acceptability, fill = buying)) +
  geom_bar(position = "dodge") +
  labs(title = "Acceptability by Buying Price",
       x = "Acceptability", y = "Count", fill = "Buying Price") +
  theme_minimal()
```

### Acceptability vs. Safety

```{r plot-safety}
ggplot(df, aes(x = acceptability, fill = safety)) +
  geom_bar(position = "dodge") +
  labs(title = "Acceptability by Safety Level",
       x = "Acceptability", y = "Count", fill = "Safety") +
  theme_minimal()
```

### Acceptability vs. Persons

```{r plot-persons}
ggplot(df, aes(x = acceptability, fill = persons)) +
  geom_bar(position = "dodge") +
  labs(title = "Acceptability by Number of Persons",
       x = "Acceptability", y = "Count", fill = "Persons") +
  theme_minimal()
```

### Acceptability vs. Maintenance Cost

```{r plot-maint}
ggplot(df, aes(x = acceptability, fill = maint)) +
  geom_bar(position = "dodge") +
  labs(title = "Acceptability by Maintenance Cost",
       x = "Acceptability", y = "Count", fill = "Maintenance") +
  theme_minimal()
```

***

# Feature Relationships

## Mutual Information Matrix

Mutual information (MI) measures the **strength** of a relationship between two variables — how much knowing one variable reduces uncertainty about the other.

> **Note:** The diagonal represents *entropy* (self-information) — how uncertain we are about a variable on its own.

```{r mutual-info}
m <- mutinformation(df)
m
```

The values most relevant for modelling are the MI scores between each feature and the target (`acceptability`):

```{r mi-target}
m[7, -7]           # MI of acceptability vs. all other features
max(m[-7, 7])      # Feature with the highest MI with acceptability
```

## Chi-Squared Tests

The chi-squared test compares:

- **Observed counts** — what you see in the data
- **Expected counts** — what you'd see if the variables were completely unrelated

A **large χ²** means observed counts diverge strongly from expected → evidence of a relationship.

### p-value Interpretation

| p-value         | Interpretation                                            |
|-----------------|-----------------------------------------------------------|
| p > 0.05        | Could easily be random luck → no significant relationship |
| p ≤ 0.05        | Unlikely to be luck → statistically significant           |
| p < 0.01        | Very unlikely → strong evidence of a relationship         |
| p < 0.001       | Vanishingly unlikely → overwhelming evidence              |

> **The Courtroom Analogy:** Defendants are presumed innocent (the *null hypothesis* H₀: no relationship). The prosecution must provide enough evidence (small p-value) to convict (reject H₀). Weak evidence (large p-value) → keep H₀.

### Buying vs. Acceptability

```{r chisq-buying-acc}
chisq.test(df$buying, df$acceptability)
```

With **p < 2.2e-16**, we have overwhelming evidence that buying price and acceptability are related. The degrees of freedom `df = 9` comes from (4 levels − 1) × (4 levels − 1) = 9.

### Buying vs. Maintenance

```{r chisq-buying-maint}
chisq.test(df$buying, df$maint)
```

### Safety vs. Acceptability

```{r chisq-safety-acc}
chisq.test(df$safety, df$acceptability)
```

> Both mutual information and chi-squared agree that **safety is strongly tied to acceptability**. When two independent methods reach the same conclusion, confidence in the finding increases — this is called *triangulation of evidence*.

***

# Train / Test Split

Splitting the data before modelling is essential to get an honest estimate of how well a model will generalise to new, unseen data.

- **Overfitting** — model performs well on training data but poorly on new data
- **Underfitting** — model is too simple to even capture the training data well

```{r split}
set.seed(7463)  # For reproducibility
split    <- sample.split(df$acceptability, SplitRatio = 0.8)
df_train <- subset(df, split == TRUE)
df_test  <- subset(df, split == FALSE)
```

## Verifying the Split

Always sanity-check after splitting to confirm class proportions are preserved.

```{r verify-split}
cat("Training rows:", nrow(df_train), "\n")
cat("Test rows:    ", nrow(df_test),  "\n")

prop.table(table(df_train$acceptability))
prop.table(table(df_test$acceptability))
```

The proportions should be very similar between train and test — confirming a stratified split.

***

# Machine Learning Algorithms

The assignment covers five classification algorithms:

| Algorithm                | Intuition                                                  |
|--------------------------|------------------------------------------------------------|
| **Decision Tree (C5.0)** | Splits data using a series of yes/no questions             |
| **Naive Bayes**          | Estimates class probabilities from feature likelihoods     |
| **Logistic Regression**  | Models class probabilities using a linear boundary         |
| **SVM**                  | Finds the maximum-margin hyperplane separating classes     |
| **Neural Network**       | Learns non-linear patterns through layered transformations |

## The Universal ML Workflow

For every algorithm, the steps are identical:

```
1. TRAIN   → model = algorithm(train_features, train_target)
2. PREDICT → predictions = predict(model, test_data)
3. EVALUATE → confusionMatrix(predictions, true_labels)
```

***

## Decision Tree (C5.0)

Decision trees build a series of if/else rules by asking the question that most reduces uncertainty at each step (measured by *information gain*).

**Why C5.0?** It is an evolution of the classic ID3 → C4.5 → C5.0 lineage, with better handling of categorical data and built-in pruning to reduce overfitting.

> **Analogy — 20 Questions:** Is safety low? → Yes → predict `unacc`. No → Is persons = 2? → Yes → predict `unacc`. No → ask another question…

### Training

```{r dt-train}
dt_model <- C5.0(df_train[-7], df_train$acceptability)
```

### Prediction

```{r dt-predict}
dt_pred <- predict(dt_model, df_test)
```

### Evaluation

A **confusion matrix** is a table that summarises how well a classification model performed by comparing its predictions against the actual known labels. Each **row** represents what the model *predicted*, and each **column** represents the *actual* class. 

- The **diagonal cells** (top-left to bottom-right) are correct predictions.
- The **off-diagonal cells** are mistakes — where the model predicted one class but the true class was different.

For example, if the model predicted `acc` but the car was actually `good`, that count appears in the `acc` row, `good` column. The further values are from the diagonal, the worse the model is performing.

```{r dt-evaluate}
confusionMatrix(dt_pred, df_test$acceptability)
```

**Key metrics to focus on:**

| Metric                         |     What it means                                                                    |
|--------------------------------|--------------------------------------------------------------------------------------|
| **Accuracy**                   | % of all predictions that were correct (96.5%)                                       |
| **Kappa**                      | Accuracy adjusted for chance — 0.925 indicates excellent agreement                   |
| **Sensitivity**                | Of all *actual* cars in a class, how many were correctly identified (e.g., `good` = 87.5%) |
| **Specificity**                | Of all cars *not* in a class, how many were correctly excluded (e.g., `acc` = 97.4%) |
| **Balanced Accuracy**          | Average of Sensitivity & Specificity — useful when classes are imbalanced            |
| **Pos Pred Value (Precision)** | When the model predicts a class, how often it is correct (e.g., when it predicts `good`, it is right 77.8% of the time) |
| **Neg Pred Value**             | When the model predicts *not* a class, how often it is correct (e.g., 99.5% for `good` — almost never misses a non-good car) |
| **Prevalence**                 | How common each class actually is in the test set (`unacc` dominates at 69.5%) |
| **Detection Rate**             | % of the entire test set correctly identified as that class — True Positives divided by total cases |
| **Detection Prevalence**       | % of the entire test set *predicted* as that class — if much higher than Detection Rate, the model is over-predicting that class |

Because `vgood` had very few cases in the test set, the model achieved 100% Sensitivity and Specificity for that class, but this should be interpreted cautiously given the small sample.

***

## Naive Bayes

Where decision trees ask *questions*, Naive Bayes calculates *probabilities*. It estimates the probability of each class given the observed features, then picks the class with the highest probability.

> **Analogy — Identifying Fruit:** You see something yellow, curved, and 15 cm long. Your brain combines past experience ("most yellow curved fruits are bananas") to conclude — with high confidence — it's a banana. Naive Bayes does exactly this using **Bayes' Theorem**.

The "naive" part: it assumes all features are *independent* of each other given the class, which is rarely true in practice but works surprisingly well.

### Training, Prediction & Evaluation

```{r nb-all}
set.seed(7463)
nb_model <- naiveBayes(df_train[, -7], df_train$acceptability)
nb_pred  <- predict(nb_model, df_test)
confusionMatrix(nb_pred, df_test$acceptability)
```

***
## Multinomial Logistic Regression

Multinomial logistic regression extends binary logistic regression to handle more than two classes. It fits one model per class against a reference class, then converts the log-odds into class probabilities using the **softmax** function.

> **Analogy — Opinion Polls:** Each feature casts a weighted vote for each class. The class that accumulates the most evidence (highest probability) wins.

The `multinom()` function from the `nnet` package is used here. `trace = FALSE` suppresses per-iteration output.

### Training, Prediction & Evaluation

```{r lg-all}
set.seed(7463)
lg_model <- multinom(acceptability ~ ., data = df_train, trace = FALSE)
lg_pred  <- predict(lg_model, df_test)
confusionMatrix(lg_pred, df_test$acceptability)
```

***

## Support Vector Machine (SVM)

SVMs find the **maximum-margin hyperplane** that best separates classes in feature space. Using the **radial basis function (RBF) kernel**, SVMs can model non-linear decision boundaries by implicitly mapping data into a higher-dimensional space.

> **Analogy — Widest Road:** Imagine drawing a road between two groups of points. SVM finds the road that is as wide as possible — the points touching the road edges are the "support vectors".

### Training, Prediction & Evaluation

```{r svm-all}
set.seed(7463)
svm_model <- ksvm(acceptability ~ ., data = df_train, kernel = "rbfdot")
svm_pred  <- predict(svm_model, df_test)
confusionMatrix(svm_pred, df_test$acceptability)
```

***

## Neural Network

Neural networks learn patterns through layered transformations. An **input layer** receives the features, one or more **hidden layers** apply non-linear activations, and an **output layer** produces class probabilities via softmax.

> **Analogy — Brain Neurons:** Just as biological neurons fire when stimulated enough, artificial neurons activate when a weighted sum of inputs exceeds a threshold — cascading through layers to produce a prediction.

The `nnet()` function fits a single hidden-layer network. `size` controls the number of hidden units; `maxit` sets the maximum training iterations.

### Training, Prediction & Evaluation

```{r nn-all}
set.seed(7463)
nn_model <- nnet(acceptability ~ ., data = df_train,
                 size = 5, maxit = 200, trace = FALSE)
nn_pred  <- predict(nn_model, df_test, type = "class")
confusionMatrix(
  factor(nn_pred, levels = levels(df_test$acceptability)),
  df_test$acceptability
)
```

***

# Model Comparison

The table below summarises overall accuracy for each model on the held-out test set.

```{r model-comparison}
results <- data.frame(
  Model = c("Decision Tree (C5.0)", "Naive Bayes",
            "Logistic Regression", "SVM", "Neural Network"),
  Accuracy = c(
    confusionMatrix(dt_pred,  df_test$acceptability)$overall["Accuracy"],
    confusionMatrix(nb_pred,  df_test$acceptability)$overall["Accuracy"],
    confusionMatrix(lg_pred,  df_test$acceptability)$overall["Accuracy"],
    confusionMatrix(svm_pred, df_test$acceptability)$overall["Accuracy"],
    confusionMatrix(
      factor(nn_pred, levels = levels(df_test$acceptability)),
      df_test$acceptability
    )$overall["Accuracy"]
  )
)

results$Accuracy <- round(results$Accuracy * 100, 2)
results[order(-results$Accuracy), ]
```

```{r plot-comparison}
ggplot(results, aes(y = reorder(Model, Accuracy), x = Accuracy)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = paste0(Accuracy, "%")), hjust = -0.1, size = 3.5) +
  labs(title = "Model Accuracy on Test Set",
       y = NULL, x = "Accuracy (%)") +
  theme_minimal() +
  xlim(0, 105)
```
